Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data

### 2022-12-13 analysis/TAS_memo/memo_TAS_reviewers.Rmd
second round of revisions to TAS memo

This memorandum provides our replies to comments provided by the TAS associated editor and two reviewers.  It indicates how we have revised the paper accordingly, as well as our understanding of the issues raised by those comments.

# Comments of Associate Editor

@@

First, as page X, lines Y of the paper indicate, the law of _Batson_ and parallel State law has consistently permitted judges to infer illegal strike bias based in part on an attorney's use of strikes in past cases. In reviewing the kinds of evidence that can prove a _Batson_ violation by a prosecutor, the U.S. Supreme Court included, as examples of types of evidence, the "relevant history of the State’s peremptory strikes in past cases." Flowers v. Mississippi, 139 S. Ct. 2228, pin (2019).

More recent State law variants on _Batson_ are also explicit about this. For example, Washington State law provides that in deciding whether "an objective observer could view race or ethnicity as a factor in the use of the peremptory challenge," the circumstances the court "should" consider include "whether the party has used peremptory challenges disproportionately against a given race or ethnicity, in the present case _or in past cases_." Wash. General Rule 37(g)(v) (emphasis added). California law similarly identifies, as a circumstance a court may consider, whether the striking "has used peremptory challenges disproportionately against a given race, ethnicity, gender, gender identity, sexual orientation, national origin, or religious affiliation, or perceived membership in any of those groups, in the present case _or in past cases_." Calif. Code of Civil Procedure 231.7(d)(3)(G) (emphasis added).

@@

Second, the phrase "should be reprimanded in the future (in some cases for no reason)" seems to be premised either on worries that our approach (1) requires trial judges to infer present illegal strike bias (i.e., find _Batson_ violations ) _solely_ based on strikes in past cases; and (2) will cause judges to be more likely to find _Batson_ violations when in fact no strike bias exists. Neither is the case, and we have revised the paper to make this more clear.

As an initial matter, the word "reprimanded" refers to a court finding of a _Batson_ violation. Our approach, however, relies on an attorney's use of strikes of past cases, regardless of whether those strikes were ever the subject of a _Batson_ challenge. This is better than just "tracking those who have been reprimanded in the past" for _Batson_ violations, because the probability of a court that finds a _Batson_ violation depends on an attorney having raised a _Batson_ challenge in the first place. Absent an attorney arguing _Batson_, a court will not inquire into whether illegal strike bias existed. If illegal strike bias exists but the attorney erroneously perceives no credible basis for it, the attorney will tend not to argue _Batson_. In cases where there no illegal strike bias but the attorney erroneously perceives some credible basis for it, the attorney will tend to argue _Batson_ but may not persuade the trial judge that illegal strike bias was more likely than not. By relying on past strike data, not whether anyone argued _Batson_ in those past cases, we avoid any selection bias associated with relying only on past court findings of _Batson_ violations.

More importantly, the law does not permit a trial judge to find illegal strike bias based _only_ on an attorney's strikes in past cases while ignoring evidence tending to prove no illegal bias in the present case. As the paper now explains on page X, line Y, _Batson_ procedure permits any type of admissible evidence that is relevant to strike bias (for or against) -- not just the attorney's past strike data -- and requires the trial judge to weigh all such evidence to decide whether illegal strike bias exists.

For this reason, our approach alone is not intended as an all-purpose detector of _Batson_ violations, because our approach cannot incorporate the weight of all the other kinds of evidence that could be submitted to prove illegal strike bias in the present case. For example, _Batson_ procedure requires the striking attorney at a certain stage to justify the challenged strikes based on (non-discriminatory) reasons. If the attorney fails to produce such reasons, the trial judge must find a _Batson_ violation. If the attorney offers such reasons, the trial judge must assess evidence as to the credibility of those reasons. Our approach cannot adjust the posterior distribution of the bias parameter to account for the credibility of those reasons or other types of evidence (other than the attorney's pattern of strikes) relevant to illegal strike bias.

Rather, as Figure 1 now depicts, our approach helps attorneys evaluate the inference of illegal bias in the present case based on an attorney's strikes in the present case and past cases, i.e., _before_ accounting for other evidence relevant to illegal strike bias in the present case (the _Batson_ violation itself). For that task, our approach assigns at most equal weight to past strikes and their strikes in the present case (i.e., when $\alpha = 1$). And if there is incompatibility betwen the historical strike data and the current strike data ($b_{hist} \neq b_{curr}$) for reasons identified on page X, lines Y, that inference of bias will be sensitive to the choice of $\alpha$, and therefore that inference of bias can and should be taken as weaker for that reason.

In turn, attorneys can use that information, _combined with_ other evidence, to (1) decide whether to argue _Batson_, (2) to prove the _Batson_ prima facie case, and (3) to persuade the trial judge to find that illegal strike bias is more likely than not, given all the evidence, i.e., find a _Batson_ violation.

@@

As explained above, the law of _Batson_ permits the inference of present strike bias based on past strikes in previous trials. The possible incompatability of $b_hist$ and $b_curr$ is thus already an issue when using past strike data to infer bias. Our contribution is to bring that issue to foreground and to advise assessing how sensitive the inference of bias is to the choice of $\alpha$ values by choosing multiple $\alpha$ values.

@@

We intend to indicate that the larger the credible interval, the greater the chance that zero falls within it.  Accordingly, the choice of the credible interval is a decision about how conservative one wishes to be about inferring bias (defined here as when $b = 0$ falls outside the chosen credible interval.

Just focus on one in the paper, rest in the appendix. It is a decision about how conservative you want to be. We have removed the discussion of the different CIs and foculs on the 95% CI for the manuscript. We note that this does not necessarily fixed the type I error, because this coverage probability refers to the true bias parameter (that is fixed in our simulations), and we are interested in "bias detection", i.e. whether the CI includes 0 or not. We also, present sensitivity analysis for the selection of the nominal HPD level. In practice, this needs to be chosen based on how conservative the lawyer bringing up a _Batson_ challenge behaves.

@@

[REPLY]

@@

We respectfully disagree with the analogy to "p-hacking", for three reasons.  First, the phrase "p-hacking" typically refers to the practice of using different statistical models on the same data in search of one or more associations that qualify "statistically significant" (p < .05), and then reporting only the analysis with the model(s) that produce those results. In contrast, in our approach, the statistical model necessarily precedes any analysis, because it has to qualify as a valid model of the strike _procedure_ of the court of interest, which can and must be known before any data is collected or analyzed.

@@

@@

Thank you for catching this.  We have revised the paper (p. X, lines Y) to clarify the values used in the the simulation for $b_{hist}$.

@@

@@

>4. In Section 3, the authors state that "the anticipated degree of incompatibility should inuence which value of $\alpha$ to select." As a follow-up of comment 2, if the assumption of compatibility is correct, setting $\alpha$ = 1 does lead to better power without sacrificing the Type I error. However, based on results in Figures 1 to 3, if this assumption is wrong, choosing a large $\alpha$ leads to either low power (bias detection rate) or large Type I error, depending on the bias in historical data. In the absence of such knowledge of compatibility, how does one choose the value of $\alpha$? In addition, to get around this issue, is it reasonable to treat $\alpha$ itself random and introduce a hyper prior?

Absent some very strong assumptions, there is no generally justifiable hyper prior for $\alpha$, because the proper weight to place on historical data depends on whether, for that historical data, there is any missing data is missing not at random and whether an inference of bias based on that historical strike data is complicated by plausible confounding or mediating trial-level variables (e.g., defendant race, criminal charge severity). The plausible reasons for non-random missingness and confounding, however, likely vary considerably depending on strike procedure, courtroom practices, prosecutor charging practices, among other things. [MORE].

@@

Reviewer 2 questions whether the data required for such a model are readily available for broad application, in addition to pointing out several important questions about the details of the paper.

@@

Review 1
Journal: The American Statistician
Manuscript ID: TAS-22-165
Title: Bayesian detection of bias in peremptory challenges using historical strike data
The authors presented a Bayesian model in detecting potential biases against minority groups
in peremptory challenges. The use of the power prior to incorporate historical data is in-
teresting, and the methodology overall seems reasonable. The proposed R Shiny App may
also be helpful to practitioners. My main concerns are about the simulation results and the
choice of the hyperparameter 
in the power prior.
Major comments:

1. In Section 2.3, the paragraph in line 47 on p. 6 ends with bhist = f3;2; : : : ; 3g,but in the very next paragraph, bhist is set to f0; 0:5; : : : ; 3g. I don't see bhist =f3;2; : : : ; 2; 3g mentioned anywhere else, including in the results shown in Figures1 to 3. I am confused about which set of values was used in the simulation for bhist.

* Fix bhist typo[DONE]

* Repeat metal detector response here.

* Ask Xiaomeng about this in the app [TEM:  reach out]

* Repeat metal detector response here. Could mention this as something that we considered, and mention that we would like find a way to measure compatibility. This is also tool intended to be used by lawyers in real-time.

Main Comments:
1. The method is first illustrated in a simulation study, with a range of values of $\alpha$.
This is fine for an audience of statisticians and attorneys specializing in jury
discrimination but is somewhat overwhelming for a general reader. It would be
easier for the reader if one data set, perhaps from the actual data from Connecticut
was used and the reader could see how the data from each additional previous trial
impacts the posterior. Perhaps only a few values of $\alpha$, rather than 11 and two intervales
of the bias parameter b, could be used.

@@

Add real data example to introduction - make more explicit

Table of bias detection for the prosecutor, when the defense shows bias. N historical = 3, nstrike =15.

	1.0

>>>>

This memorandum replies to comments provided by the TAS associate editor and two reviewers.  It indicates how we have revised the paper accordingly, as well as our understanding of the issues raised by those comments.

# Comments of TAS Associate Editor

@@

First, as page X, lines Y of the paper indicate, the law of _Batson_ and parallel State law has consistently permitted judges to infer illegal strike bias based in part on an attorney's use of strikes in past cases. In reviewing the kinds of evidence that can prove a _Batson_ violation by a prosecutor, the U.S. Supreme Court included, as examples of types of evidence, the "relevant history of the State’s peremptory strikes in past cases." Flowers v. Mississippi, 139 S. Ct. 2228, pin (2019). More recent State law variants on _Batson_ are also explicit about this. For example, Washington State law provides that in deciding whether "an objective observer could view race or ethnicity as a factor in the use of the peremptory challenge," the circumstances the court "should" consider include "whether the party has used peremptory challenges disproportionately against a given race or ethnicity, in the present case _or in past cases_." Wash. General Rule 37(g)(v) (emphasis added). California law similarly identifies, as a circumstance a court may consider, whether the striking "has used peremptory challenges disproportionately against a given race, ethnicity, gender, gender identity, sexual orientation, national origin, or religious affiliation, or perceived membership in any of those groups, in the present case _or in past cases_." Calif. Code of Civil Procedure 231.7(d)(3)(G) (emphasis added).

@@

Second, as the phrase "should be reprimanded in the future (in some cases for no reason)" implies, the associate editor seems worried that our approach will cause judges to infer that illegal strike bias is more likely than not (i.e., find _Batson_ violations ) when in fact no strike bias exists.

This is not the case, and we have revised the paper to make this more clear.

First, we assume the word "reprimanded" refers to a court finding of a _Batson_ violation. Our approach, however, relies on an attorney's use of strikes of past cases, regardless of whether those strikes were ever the subject of a _Batson_ challenge. This is better than just "tracking those who have been reprimanded in the past" for _Batson_ violations, because the probability of a court that finds a _Batson_ violation depends on an attorney having raised a _Batson_ challenge in the first place. (Absent an attorney arguing _Batson_, a court will not inquire into whether illegal strike bias existed.) If illegal strike bias exists but the attorney erroneously perceives no credible basis for it, the attorney will tend not to argue _Batson_. In cases where there no illegal strike bias but the attorney erroneously perceives some credible basis for it, the attorney will tend to argue _Batson_ but may not persuade the trial judge that illegal strike bias was more likely than not. By relying on past strike data, not whether anyone argued _Batson_ in those past cases, we avoid any selection bias associated with relying only on past court findings of _Batson_ violations.

Second, _Batson_ procedure permits any type of admissible evidence of strike bias (for or against) -- not just the attorney's past strike data -- and requires the trial judge to weigh all such evidence to decide whether illegal strike bias exists. For this reason, our approach alone is not an all-purpose detector of _Batson_ violations, because our approach cannot incorporate the weight of all the other kinds of evidence that could be submitted to prove illegal strike bias in the present case. For example, _Batson_ procedure requires the striking attorney at a certain stage to justify the challenged strikes based on (non-discriminatory) reasons. If the attorney fails to produce such reasons, the trial judge must find a _Batson_ violation. If the attorney offers such reasons, the trial judge must assess evidence as to the credibility of those reasons.

Our approach cannot adjust the posterior distribution of the bias parameter to account for the credibility of those reasons or other types of evidence (other than the attorney's pattern of strikes) of illegal strike bias. Rather, as the new Figure 1 depicts, our approach helps attorneys evaluate the inference of illegal bias in the present case based on historical strike data but _before_ accounting for other evidence relevant to illegal strike bias. For that task, our approach assigns at most equal weight to past strikes and their strikes in the present case (i.e., when $\alpha = 1$).

The possible incompatability of $b_hist$ and $b_curr$ is already and always an issue when using past strike data to infer bias. This is a reality faced by _any_ statistical technique for inferring strike bias using past strike data. 
Because the law of _Batson_ expressly permits inferring current strike bias based on past strikes in previous trials, the lawyers in _Batson_ challenges involving past strike data are already and always arguing about how much weight to assign to that past strike data.

Our approach simply makes the weighting of historical strike data -- usually expressed with words -- more explicit and transparent by requiring a choice of a value for alpha. The more sensitive the inference of bias (zero falls outside the credible interval) is to the choice of $\alpha$, the more one should worry about incompatibility. In turn, the appropriateness of small vs. large values of $\alpha$ vary with the suspected degree of incompatability. In turn, attorneys can use that information, _combined with_ other evidence, to (1) decide whether to argue _Batson_, (2) to prove the _Batson_ prima facie case, and (3) to persuade the trial judge to find that illegal strike bias is more likely than not, given all the evidence, i.e., find a _Batson_ violation.

We hope that these modifications clarify the placement of our model in the steps from a sequence of peremptory challenges to a _Batson_ challenge actually being brought, and upheld by the trial judge.

@@

Finally, the reviewer's use of the word "guilty" requires two clarifying notes.  First, _Batson_ violations must be proven under the "preponderance of the evidence" standard (more likely than not), _not_ the more stringent "reasonable doubt" standard that applies when a prosecutor aims to prove a defendant "guilty" of committing the criminal offense charged. Second, as explained above, bias "detection" in the paper means only whether the credible interval for the bias parameter ($b_{curr}) excludes zero. Because _Batson_ procedure requires weighing any and all evidence of strike bias, not just past strike data, our approach cannot alone identify _Batson_ violations in cases involving other relevant evidence bias besides past strike data.

@@

We have revised this discussion to remove such confusion.  The choice of the credible interval is a decision about how conservative one wishes to be about inferring bias from past strike data alone. We define bias "detection" narrowly to mean when $b = 0$ falls outside the chosen credible interval. Thus, the larger the credible interval, the greater the chance that zero falls within it. For the sake of clarity, we now focus on one credible interval (95%) in the paper and have placed simulation results based on other credible intervals in the Appendix. This does not necessarily fix the type I error, because this coverage probability refers to the true bias parameter (that is fixed in our simulations), and we are interested in "bias detection", i.e. whether the credible interval includes 0 or not. We also present sensitivity analysis for the selection of the nominal HPD level.

@@

This comment is not clear to us. The real data analyses are based on real _historical_ trial data from the federal district court in Connecticut (see lines x to y). Users of the app have the option to run the model excluding any historical trial data (as suggested by the comment) by not selecting the name of a prosecutor or defense attorney. In this case, the normal(0,2) prior is used for the bias parameter. Similarly, the app gives the user the option of altering the weight the historical strike data (if selected) by adjusting alpha (power prior parameter). Thus the user can: 1) choose not to use historical data, 2) use historical data with equal weight as the current trial (alpha = 1), or 3) down-weight the historical trial data.

@@

We respectfully disagree with the analogy to "p-hacking", for two reasons.  First, the phrase "p-hacking" typically refers to the practice of using different statistical models on the same data in search of one or more associations that qualify as "statistically significant" (p < .05), and then reporting only the analysis with the model(s) that produce those results. In contrast, in our approach, the statistical model necessarily precedes any analysis, because it has to qualify as a valid model of the strike _procedure_ of the court of interest, which can and must be known before any data is collected or analyzed.

@@

[INSERT] Batson challenge doesn't care if you strike a cognizable class because you assume they have particular confounders (e.g., low income, etc). [editme] We do explicitly highlight that trial-level characteristics and impact compatibility between current and historical trials. We do advocate for reassessment of the way (and what) data are collected from individual trials.

@@

Thank you for catching this.  We have revised the paper (p. X, lines Y) to clarify the values used in the simulation for $b_{hist}$.

@@

We agree that the bias "detection" rate (whether the credible interval excludes zero) can be as high as 100% when true bias is zero. We also agree that the power to detect bias is low when there is low bias in historical trials. These situations can occur because of high incompatibility.  Similarly, the low power situations occurs due the limited number of strikes, which is why _compatible_ historical strike data can increase this power.

This is an advantage of using the power prior: We can adjust the weight of the historical strike data (and therefore type I error) by adjusting alpha. In reality, because $b_{curr}$ and $b_{hist}$ are not directly observable, one will need to rely on background knowledge or assumptions about the degree of incompatibility to decide which value of alpha to choose. If they bring a _Batson_ challenge, they must also disclose and justify their choice of alpha and the assumptions on which that choice is based.

@@

[INSERT] Will get Eric to check this (still no response from Xiaomeng)

>4. In Section 3, the authors state that "the anticipated degree of incompatibility should influence which value of $\alpha$ to select." As a follow-up of comment 2, if the assumption of compatibility is correct, setting $\alpha$ = 1 does lead to better power without sacrificing the Type I error. However, based on results in Figures 1 to 3, if this assumption is wrong, choosing a large $\alpha$ leads to either low power (bias detection rate) or large Type I error, depending on the bias in historical data. In the absence of such knowledge of compatibility, how does one choose the value of $\alpha$? In addition, to get around this issue, is it reasonable to treat $\alpha$ itself random and introduce a hyper prior?

As indicated in the response to comment 2 above, one chooses a value of $alpha$ based on background beliefs on plausible degree of incompatibility, and then evaluate how sensitive bias detection (whether zero falls outside the credible interval) is to the choice of alpha value. Absent some very strong assumptions, there is no generally justifiable hyper prior for $\alpha$, because the proper weight to place on historical data depends on whether, for that historical data, there is any missing data that is missing not at random and whether an inference of bias based on that historical strike data is complicated by plausible confounding or mediating trial-level variables (e.g., defendant race, criminal charge severity). The plausible reasons for non-random missingness and confounding, however, likely vary considerably depending on strike procedure, courtroom practices, prosecutor charging practices, among other things.

@@

>1. The method is first illustrated in a simulation study, with a range of values of $\alpha$. This is fine for an audience of statisticians and attorneys specializing in jury discrimination but is somewhat overwhelming for a general reader. It would be easier for the reader if one data set, perhaps from the actual data from Connecticut was used and the reader could see how the data from each additional previous trial impacts the posterior. Perhaps only a few values of $\alpha$, rather than 11 and two val[u]es of the bias parameter b, could be used.

In the Discussion, we explained that data on consecutive strike patterns is hard to collect, simply because of the challenges of collecting any historical strike data, regardless of strike procedure. We have revised that Discussion to add that, since the 1980s, any person has a First Amendment right to attend jury selection in criminal court proceedings, and criminal defendants have a Sixth Amendment right to public court proceedings during jury selection. These legal rights, in turn, enable data collection by court observers, regardless of State and local recordkeeping practices.

Finally, the current availability of the data required by the model seems orthogonal to the merits of our statistical approach of using such data to infer bias. [MORE].

@@

Main Comments:

@@
